---
subtitle: "STAT 3480 Consulting Project"
title: "Coronavirus in the Southeast"
author: "Ishaan Dey & Tal Dunne"
date: "11/23/2020"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=F, message=F)
```
*****
# Executive Summary


\newpage
# Introduction
## Project Description
- Restate the client’s research problem (for clarification)
- Put the data and the research question in context.
- Why is the client’s research relevant?
- Has this research question been studied previously?

## Data Summary & Discussion
- How was the data collected?
– What is the sampling/ experimental design?
– What are some potential biases?
- What variables are measured? What are the units?
- How large is the dataset?
- Some basic summary statistics and graphs of the data should be presented. (These basic distributions will be relevant for the methodology.)

\newpage
# Methodology
- What variables are measured? What are the units?
- How large is the dataset?
- Some basic summary statistics and graphs of the data should be presented. (These basic distributions will be relevant for the methodology.)

# Results
- Test statistics
- P-values
- Confidence intervals
- Inference decisions (conclusions

# Discussion
- Interpret and discuss the significance of the analysis results.
- Substantiate any conclusions with results.
- Tell the story of how you answered your client’s research question(s).
- Discuss the limitations of the study (which can pertain to data, data collection, or methodology).
- Make sure to be delicate – the client doesn’t want to hear what a bad job they’ve done.

# Conclusion
- The goal of this section is to address the client’s bottom line: what does your progress mean for his/her research problem?

\newpage
# Appendix
## Code

```{r library, warning=F, results='hide'}
library(readr)
library(dplyr)
```

```{r import, warning=F}
covid_data <- readr::read_csv("covid-data.csv") %>% as_tibble()
covid_data$Mask.Mandate <- dplyr::if_else(covid_data$Mask.Mandate.Start.Date == 'None', 
                                          true=0, false=1) %>% as.factor()
covid_data %>% head()
```
### Big Q1
```{r q1-setup}
mask.sample1 <- covid_data %>% 
  filter(Mask.Mandate == 1) %>% pull(Cases.Per.1000)
mask.sample2 <- covid_data %>% 
  filter(Mask.Mandate == 0) %>% pull(Cases.Per.1000)
```

```{r q1-test, results=T}
wilcox.test(mask.sample1, mask.sample2, alternative="less")
```

### Big Q2
```{r}
case.rate <- covid_data$Cases.Per.1000  
death.rate <- covid_data$Death.Rate.Pct  
```

```{r q2-test-spearman}
r.obs<-cor(rank(case.rate), rank(death.rate))

source("http://www4.stat.ncsu.edu/~lu/ST505/Rcode/functions-Ch5.R")
perm.r<-perm.approx.r(rank(case.rate),rank(death.rate),2000)

p.upper<-mean(perm.r >= r.obs)
p.lower<-mean(perm.r <= r.obs)
p.two.sided<-mean(abs(perm.r) >= abs(r.obs))

paste("Spearman's Correlation:", round(r.obs,4)) %>% print()
paste("p-value:", round(p.upper,3)) %>% print()
```

```{r}
tau.obs<-getTau(case.rate,death.rate)

perm.tau<-perm.approx.tau(case.rate,death.rate,2000)

p.upper.tau<-mean(perm.tau >= tau.obs)
p.lower.tau<-mean(perm.tau <= tau.obs)
p.two.sided.tau<-mean(abs(perm.tau) >= abs(tau.obs))

paste("Kendall's Tau:", round(tau.obs,4)) %>% print()
paste("p-value:", round(p.upper.tau,3)) %>% print()
```


## Figures
*Figure 1*. Distribution of Case Rates (per 1000), by State Mask Mandates.

```{r q1-eda, echo=F}
par(mfrow=c(2,1))
breaks <- seq(10, 40, 2)
hist(mask.sample1, xlab='', main='Mask Mandated', 
     ylab='No. of States', breaks = breaks)
hist(mask.sample2, xlab='Case Rate Per 1000', main='Mask Not Mandated',
     ylab='No. of States',breaks = breaks)
```


*Figure 2*. Scatterplot of Death Rate (%) against Case Rates (per 1000) 

```{r}
plot(case.rate, death.rate, main='Death Rate (%) vs. Case Rates (per 1000)',
     xlab='Case Rates (per 1000)', ylab = 'Death Rate (%)')
```

